---
title: "Life Expectancy"
author: "Rick Fontenot"
date: "5/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Libraries
```{r load-packages, include=FALSE}
library(dplyr)
library(tidyverse)
library(caret)
library(DataExplorer)
library(gplots)
library(graphics)
library(corrplot)
library(olsrr)
library(ggpubr)
library(rstatix)
library(dplyr)
library(tidyverse)
library(visdat)
library(GGally)
library(usmap)
library(mice)
library(VIM)
library(plotly)
library(caret)
library(e1071)
library(class)
library(maps)
library(mapproj)
library(stringr)
library(ggplot2) 
library(ggthemes)
library(table1)
library(DataExplorer)
```
Load Theme for plots

```{r}
theme_set(theme_fivethirtyeight())
theme_update(axis.title = element_text()) #the default for fivethirtyeight is to not show axis labels, this removes that default so we can choose to specify and display axis titles
theme_update(plot.title = element_text(hjust = 0.5)) # changing default to center all titles
```

Load Data from Kaggle

From https://www.kaggle.com/philbowman212/life-expectancy-exploratory-data-analysis

Variable Descriptions
Format: variable (type) - description

country (Nominal) - the country in which the indicators are from (i.e. United States of America or Congo)

year (Ordinal) - the calendar year the indicators are from (ranging from 2000 to 2015)

status (Nominal) - whether a country is considered to be 'Developing' or 'Developed' by WHO standards

life_expectancy (Ratio) - the life expectancy of people in years for a particular country and year

adult_mortality (Ratio) - the adult mortality rate per 1000 population (i.e. number of people dying between 15 and 60 years per 1000 population); if the rate is 263 then that means 263 people will die out of 1000 between the ages of 15 and 60; another way to think of this is that the chance an individual will die between 15 and 60 is 26.3%

infant_deaths (Ratio) - number of infant deaths per 1000 population; similar to above, but for infants

alcohol (Ratio) - a country's alcohol consumption rate measured as liters of pure alcohol consumption per capita

percentage_expenditure (Ratio) - expenditure on health as a percentage of Gross Domestic Product (gdp)

hepatitis_b (Ratio) - number of 1 year olds with Hepatitis B immunization over all 1 year olds in population

measles (Ratio) - number of reported Measles cases per 1000 population

bmi (Interval/Ordinal) - average Body Mass Index (BMI) of a country's total population

under-five_deaths (Ratio) - number of people under the age of five deaths per 1000 population

polio (Ratio) - number of 1 year olds with Polio immunization over the number of all 1 year olds in population

total_expenditure (Ratio) - government expenditure on health as a percentage of total government expenditure

diphtheria (Ratio) - Diphtheria tetanus toxoid and pertussis (DTP3) immunization rate of 1 year olds

hiv/aids (Ratio) - deaths per 1000 live births caused by HIV/AIDS for people under 5; number of people under 5 who die due to 

HIV/AIDS per 1000 births

gdp (Ratio) - Gross Domestic Product per capita

population (Ratio) - population of a country

thinness_1-19_years (Ratio) - rate of thinness among people aged 10-19 (Note: variable should be renamed to thinness_10-19_years 
to more accurately represent the variable)

thinness_5-9_years (Ratio) - rate of thinness among people aged 5-9

income_composition_of_resources (Ratio) - Human Development Index in terms of income composition of resources (index ranging from 0 to 1)

schooling (Ratio) - average number of years of schooling of a population

```{r load data}
kaggle.training = read.csv("https://raw.githubusercontent.com/JosephLazarus/Life_Expectancy/main/WHO_Data/Life%20Expectancy%20Data.csv", header = TRUE)

head(kaggle.training)

str(kaggle.training)
#Note that Country & Status are categorical, all other predictors are numerical
```

Investigate NA values to determine what needs resolution

```{r}
aggr_plot <- aggr(kaggle.training, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(kaggle.training), cex.axis=.7, gap=3, ylab=c("Percent data missing","Combinations Missing"))

#Lots of missing values to deal with >15% for Population, Hepatitis.B and GDP


marginplot(kaggle.training[c(5,4)])
#view(kaggle.training)
#

plot_histogram(kaggle.training)


```

Look for correlations to Population to explore methods of imputing missing values

```{r}
Population<-kaggle.training %>% select(-Country,-Status) %>% na.omit() %>% filter(Population<1e8)

stat.test <- cor(Population[-16], Population$Population) 

Population %>% ggplot(aes(x = infant.deaths, y = Population)) + geom_point(size=0.5)

model <- lm(Population ~ ., data = Population)
summary(model)
#Adj R=0.48 , significant variables are infant.deaths, Measles, under.five.deaths, Schooling

model2 <- lm(Population ~ infant.deaths + Measles + under.five.deaths + Schooling, data = Population)
summary(model2)
#Adj R=0.48 , significant variables are infant.deaths, Measles, under.five.deaths, Schooling

Population["prediction"]<-predict(model2, newdata = Population)

Population %>% ggplot(aes(x = Population, y = prediction)) + geom_point(size=0.5)

Population %>% ggplot(aes(x = Population, y = prediction)) + geom_point(size=0.5) + xlim(0, 1e5) + ylim(0,1e7)

#this regression is not a very good predictor

missingPopulation<-kaggle.training %>% group_by(Country) %>% summarise(MissingYears = sum(is.na(Population)))
missingPopulation<-missingPopulation[order(missingPopulation$MissingYears,decreasing=TRUE),]
missingPopulation

#40 Countries are missing population for all 16 years, these account for 640 of the 652 missing values, range years 2000-2015

#Import population data from http://wdi.worldbank.org/table/2.1#
#Note used 2000 Actual Population plus growth rate to estimate 2001 through 2015

worldbank = read.csv("https://raw.githubusercontent.com/JosephLazarus/Life_Expectancy/main/Data_Folder/WorldBankPopulation.csv", header = TRUE, fileEncoding="UTF-8-BOM")
view(worldbank)

worldbank <- worldbank %>% select(-Population2000,-Population2019,-PopulationGrowth) %>% gather(-Country, key = "Year", value = "EstPopulation")

worldbank$Year <- gsub("X", "", worldbank$Year)
worldbank$Year <- as.numeric(worldbank$Year)
worldbank$EstPopulation <- as.numeric(worldbank$EstPopulation)
worldbank

worldbank$Country <- gsub("Bahamas, The", "Bahamas", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Bolivia", "Bolivia (Plurinational State of)", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Congo, Rep.", "Congo", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Cote d'Ivoire", "Côte d'Ivoire", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Czech Republic", "Czechia", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Korea, Dem. People's Rep.", "Democratic People's Republic of Korea", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Congo, Dem. Rep.", "Democratic Republic of the Congo", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Egypt, Arab Rep.", "Egypt", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Gambia, The", "Gambia", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Iran, Islamic Rep.", "Iran (Islamic Republic of)", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Kyrgyz Republic", "Kyrgyzstan", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Lao PDR", "Lao People's Democratic Republic", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Micronesia, Fed. Sts.", "Micronesia (Federated States of)", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Korea, Rep.", "Republic of Korea", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Moldova", "Republic of Moldova", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("St. Lucia", "Saint Lucia", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("St. Vincent and the Grenadines", "Saint Vincent and the Grenadines", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Slovak Republic", "Slovakia", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Eswatini", "Swaziland", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("North Macedonia", "The former Yugoslav republic of Macedonia", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("United Kingdom", "United Kingdom of Great Britain and Northern Ireland", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Tanzania", "United Republic of Tanzania", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("United States", "United States of America", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Venezuela, RB", "Venezuela (Bolivarian Republic of)", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Vietnam", "Viet Nam", worldbank$Country) #rename to match kaggle set
worldbank$Country <- gsub("Yemen, Rep.", "Yemen", worldbank$Country) #rename to match kaggle set

kaggle.training %>% left_join(worldbank)

kaggle.training<-left_join(kaggle.training, worldbank, by = c("Country","Year"), copy = FALSE)

kaggle.training %>% ggplot(aes(x = EstPopulation, y = Population)) + geom_point(size=0.5)+geom_jitter()+ xlim(0, 1e7) + ylim(0,1e7)

#Note there are some where correlation is way off, look at Afgahnistan as example
Afghanistan <- kaggle.training %>% filter(Country=="Afghanistan") %>% select(Year,Population,EstPopulation) %>% gather(-Year, key = "Type", value = "Population")
ggplot(Afghanistan,aes(x = Year, y = Population,color=Type)) + geom_line()
#Kaggle data set appears to have error with anual population bouncing up and down drastically. Some years are off by factor of 10 some years are off by factor of 100

#Check Albania as second example
Albania <- kaggle.training %>% filter(Country=="Albania") %>% select(Year,Population,EstPopulation) %>% gather(-Year, key = "Type", value = "Population")
ggplot(Albania,aes(x = Year, y = Population,color=Type)) + geom_line()
#Albania has similar issue including some years near zero population

#Check Rwanda as second example
Rwanda <- kaggle.training %>% filter(Country=="Rwanda") %>% select(Year,Population,EstPopulation) %>% gather(-Year, key = "Type", value = "Population")
ggplot(Rwanda,aes(x = Year, y = Population,color=Type)) + geom_line()
#Rwanda has similar issue including some years near zero population

#Attempt to correct Afghanistan population errors by factors of 10 & 100 for comparison
Afghanistan <- kaggle.training %>% filter(Country=="Afghanistan") %>% select(Year,Population,EstPopulation)
Afghanistan["xPopulation"]<-Afghanistan$Population
Afghanistan[2,]$xPopulation<-Afghanistan[2,]$Population*100
Afghanistan[4,]$xPopulation<-Afghanistan[4,]$Population*10
Afghanistan[5,]$xPopulation<-Afghanistan[5,]$Population*10
Afghanistan[6,]$xPopulation<-Afghanistan[6,]$Population*10
Afghanistan[7,]$xPopulation<-Afghanistan[7,]$Population*100
Afghanistan[8,]$xPopulation<-Afghanistan[8,]$Population*10
Afghanistan[10,]$xPopulation<-Afghanistan[10,]$Population*10
Afghanistan[11,]$xPopulation<-Afghanistan[11,]$Population*100
Afghanistan[13,]$xPopulation<-Afghanistan[13,]$Population*10
Afghanistan[15,]$xPopulation<-Afghanistan[15,]$Population*10
Afghanistan[16,]$xPopulation<-Afghanistan[16,]$Population*100
Afghanistan<-Afghanistan%>% gather(-Year, key = "Type", value = "Population")
ggplot(Afghanistan,aes(x = Year, y = Population,color=Type)) + geom_line()



#Kaggle Population data doesn't look reliable, may want to use world bank instead
view(kaggle.training)

#Re-check missing data for Worldbank Population

missingPopulation2<-kaggle.training %>% group_by(Country) %>% summarise(MissingYears = sum(is.na(EstPopulation)))
missingPopulation2<-missingPopulation2[order(missingPopulation2$MissingYears,decreasing=TRUE),]
missingPopulation2
#Only 4 NA's when using EstPopulation from WorldBank data


```

Do the same analysis with GDP data from World Bank

```{r}

#Check missing GDPdata for for Kaggle set by country
missingGDP<-kaggle.training %>% group_by(Country) %>% summarise(MissingYears = sum(is.na(GDP)))
missingGDP<-missingGDP[order(missingGDP$MissingYears,decreasing=TRUE),]
missingGDP
#25 countries are missing data for all 16 years, these account for 400 of the 448 missing values. Would be tought to impute on means, look at alt source instead

gdp = read.csv("https://raw.githubusercontent.com/JosephLazarus/Life_Expectancy/main/Data_Folder/WorldBankGDP.csv", header = TRUE, fileEncoding="UTF-8-BOM")

gdp <- gdp %>% gather(-Country, key = "Year", value = "EstGDP")

gdp$Year <- gsub("X", "", gdp$Year)
gdp$Year <- as.numeric(gdp$Year)
gdp$EstGDP <- as.numeric(gdp$EstGDP)
gdp

gdp$Country <- gsub("Bahamas, The", "Bahamas", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Bolivia", "Bolivia (Plurinational State of)", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Congo, Rep.", "Congo", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Cote d'Ivoire", "Côte d'Ivoire", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Czech Republic", "Czechia", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Korea, Dem. People's Rep.", "Democratic People's Republic of Korea", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Congo, Dem. Rep.", "Democratic Republic of the Congo", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Egypt, Arab Rep.", "Egypt", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Gambia, The", "Gambia", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Iran, Islamic Rep.", "Iran (Islamic Republic of)", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Kyrgyz Republic", "Kyrgyzstan", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Lao PDR", "Lao People's Democratic Republic", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Micronesia, Fed. Sts.", "Micronesia (Federated States of)", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Korea, Rep.", "Republic of Korea", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Moldova", "Republic of Moldova", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("St. Lucia", "Saint Lucia", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("St. Vincent and the Grenadines", "Saint Vincent and the Grenadines", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Slovak Republic", "Slovakia", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Eswatini", "Swaziland", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("North Macedonia", "The former Yugoslav republic of Macedonia", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("United Kingdom", "United Kingdom of Great Britain and Northern Ireland", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Tanzania", "United Republic of Tanzania", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("United States", "United States of America", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Venezuela, RB", "Venezuela (Bolivarian Republic of)", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Vietnam", "Viet Nam", gdp$Country) #rename to match kaggle set
gdp$Country <- gsub("Yemen, Rep.", "Yemen", gdp$Country) #rename to match kaggle set

kaggle.training<-left_join(kaggle.training, gdp, by = c("Country","Year"), copy = FALSE)

#now run
kaggle.training %>% ggplot(aes(x = EstGDP, y = GDP)) + geom_point(size=0.5)+geom_jitter()

#Note there are some where correlation is way off, look at Afgahnistan as example
Afghanistan <- kaggle.training %>% filter(Country=="Afghanistan") %>% select(Year,GDP,EstGDP) %>% gather(-Year, key = "Type", value = "GDP")
ggplot(Afghanistan,aes(x = Year, y = GDP,color=Type)) + geom_line()
#Kaggle data set appears to have error with anual population bouncing up and down drastically. Some years are off by factor of 10 some years are off by factor of 100

#Check Albania as second example
Albania <- kaggle.training %>% filter(Country=="Albania") %>% select(Year,GDP,EstGDP) %>% gather(-Year, key = "Type", value = "GDP")
ggplot(Albania,aes(x = Year, y = GDP,color=Type)) + geom_line()
#Albania has similar issue including some years near zero population

#Check Rwanda as second example
Rwanda <- kaggle.training %>% filter(Country=="Rwanda") %>% select(Year,GDP,EstGDP) %>% gather(-Year, key = "Type", value = "GDP")
ggplot(Rwanda,aes(x = Year, y = GDP,color=Type)) + geom_line()
#Rwanda has similar issue including some years near zero population


#Re-check missing data for Worldbank Population
missingGDP2<-kaggle.training %>% group_by(Country) %>% summarise(MissingYears = sum(is.na(EstGDP)))
missingGDP2<-missingGDP2[order(missingGDP2$MissingYears,decreasing=TRUE),]
missingGDP2
#Democratic Rep of Korea not in world bank set and Somlia missing 2000-2012, South Sudan missing 2000-2005 Syrian Arab Rep missing 2008-2015in world bank, not a merge issue. These 4 countries account for 45 of remaining 60 missing values, but big improvement from original 448 missing values

#Bigger issue is why values in both sets don't correlate, is there a scaling issue?

view(kaggle.training)

#Data Description said kaggle GDP data was per capita, world bank is in $billions, calculate estGDP per capita
kaggle.training["EstGDPpercapita"]<-kaggle.training$EstGDP * 1000000000/ kaggle.training$EstPopulation /1000
kaggle.training %>% ggplot(aes(x = EstGDPpercapita, y = GDP)) + geom_point(size=0.5)+geom_jitter()
#scales match now but two distict relationship lines, what's going on there?

#Note there are some where correlation is way off, look at Afgahnistan as example
Afghanistan <- kaggle.training %>% filter(Country=="Afghanistan") %>% select(Year,GDP,EstGDPpercapita) %>% gather(-Year, key = "Type", value = "GDP")
ggplot(Afghanistan,aes(x = Year, y = GDP,color=Type)) + geom_line()
#Kaggle data set appears to have error with annual population bouncing up and down drastically. Some years are off by factor of 10 some years are off by factor of 100

#Note there are some where correlation is way off, look at Afgahnistan as example
Albania <- kaggle.training %>% filter(Country=="Albania") %>% select(Year,GDP,EstGDPpercapita) %>% gather(-Year, key = "Type", value = "GDP")
ggplot(Albania,aes(x = Year, y = GDP,color=Type)) + geom_line()
#Kaggle data set appears to have error with annual population bouncing up and down drastically. Some years are off by factor of 10 some years are off by factor of 100

#Check Rwanda as second example
Rwanda <- kaggle.training %>% filter(Country=="Rwanda") %>% select(Year,GDP,EstGDPpercapita) %>% gather(-Year, key = "Type", value = "GDP")
ggplot(Rwanda,aes(x = Year, y = GDP,color=Type)) + geom_line()
#Rwanda has similar issue

#These plots show same issue, GDP per capita probably confounded by same issue with bad population data

view(kaggle.training)

```
Since Population data issue caused per capita calculation issue on GDP, does it affect any otherper capita variables like:

alcohol (Ratio) - a country's alcohol consumption rate measured as liters of pure alcohol consumption per capita

percentage_expenditure (Ratio) - expenditure on health as a percentage of Gross Domestic Product (gdp)

measles (Ratio) - number of reported Measles cases per 1000 population

under-five_deaths (Ratio) - number of people under the age of five deaths per 1000 population

total_expenditure (Ratio) - government expenditure on health as a percentage of total government expenditure

```{r}
#Afghanistan had population issues for years 2005 & 2011
#Albania had population issues for years 2003, 2005, 2006,2007,2008, 2010
#Rwanda had population issues for years 2001,2007,2008,2014,2015

#Check Alcohol
Afghanistan <- kaggle.training %>% filter(Country=="Afghanistan") %>% select(Year,Alcohol) %>% gather(-Year, key = "Type", value = "Alcohol")
ggplot(Afghanistan,aes(x = Year, y = Alcohol,color=Type)) + geom_line()
#Alcohol looks OK

#Check percentage.expenditure
Afghanistan <- kaggle.training %>% filter(Country=="Afghanistan") %>% select(Year,percentage.expenditure) %>% gather(-Year, key = "Type", value = "percentage.expenditure")
ggplot(Afghanistan,aes(x = Year, y = percentage.expenditure,color=Type)) + geom_line()
#percentage.expenditure shows issue in years with population problem matches 2005,2011

Albania <- kaggle.training %>% filter(Country=="Albania") %>% select(Year,percentage.expenditure) %>% gather(-Year, key = "Type", value = "percentage.expenditure")
ggplot(Albania,aes(x = Year, y = percentage.expenditure,color=Type)) + geom_line()
#percentage.expenditure shows issue in years with population problem matches 2003, 2005,2006,2007,2008,2010

kaggle.training["CorrectedExpenditure"]<-kaggle.training$percentage.expenditure * (kaggle.training$EstGDPpercapita / kaggle.training$GDP)

Afghanistan <- kaggle.training %>% filter(Country=="Afghanistan") %>% select(Year,CorrectedExpenditure,percentage.expenditure) %>% gather(-Year, key = "Type", value = "CorrectedExpenditure")
ggplot(Afghanistan,aes(x = Year, y = CorrectedExpenditure,color=Type)) + geom_line()
#percentage.expenditure shows issue in years with population problem matches 2005,2011

Albania <- kaggle.training %>% filter(Country=="Albania") %>% select(Year,CorrectedExpenditure,percentage.expenditure) %>% gather(-Year, key = "Type", value = "CorrectedExpenditure")
ggplot(Albania,aes(x = Year, y = CorrectedExpenditure,color=Type)) + geom_line()
#percentage.expenditure shows issue in years with population problem matches 2003, 2005,2006,2007,2008,2010

#Based on these plots use Corrected Expenditure instead of raw percentage.expenditure

#Check Measles
Afghanistan <- kaggle.training %>% filter(Country=="Afghanistan") %>% select(Year,Measles) %>% gather(-Year, key = "Type", value = "Measles")
ggplot(Afghanistan,aes(x = Year, y = Measles,color=Type)) + geom_line()

Albania <- kaggle.training %>% filter(Country=="Albania") %>% select(Year,Measles) %>% gather(-Year, key = "Type", value = "Measles")
ggplot(Albania,aes(x = Year, y = Measles,color=Type)) + geom_line()

Rwanda <- kaggle.training %>% filter(Country=="Rwanda") %>% select(Year,Measles) %>% gather(-Year, key = "Type", value = "Measles")
ggplot(Rwanda,aes(x = Year, y = Measles,color=Type)) + geom_line()
#Measles doesn't appear to have population error as divisor

#Check under.five.deaths
Afghanistan <- kaggle.training %>% filter(Country=="Afghanistan") %>% select(Year,under.five.deaths) %>% gather(-Year, key = "Type", value = "under.five.deaths")
ggplot(Afghanistan,aes(x = Year, y = under.five.deaths,color=Type)) + geom_line()

Albania <- kaggle.training %>% filter(Country=="Albania") %>% select(Year,under.five.deaths) %>% gather(-Year, key = "Type", value = "under.five.deaths")
ggplot(Albania,aes(x = Year, y = under.five.deaths,color=Type)) + geom_line()
#under.five.deaths does not appear to have population error as divisor

#Check Total.expenditure
Afghanistan <- kaggle.training %>% filter(Country=="Afghanistan") %>% select(Year,Total.expenditure) %>% gather(-Year, key = "Type", value = "Total.expenditure")
ggplot(Afghanistan,aes(x = Year, y = Total.expenditure,color=Type)) + geom_line()

Albania <- kaggle.training %>% filter(Country=="Albania") %>% select(Year,Total.expenditure) %>% gather(-Year, key = "Type", value = "Total.expenditure")
ggplot(Albania,aes(x = Year, y = Total.expenditure,color=Type)) + geom_line()
#Total.Expenditure does not appear to have population error as divisor

#In review of this section all these variables checked for population error issue are ok except percentage.expenditure which we should replace with Corrected.expenditure
```

Explore regression to impute Hepatitis

```{r}

#Check missing GDPdata for for Kaggle set by country
kaggle.training %>% summarise(sum(is.na(Hepatitis.B)))
#553 missing values total
missingHepatitis<-kaggle.training %>% group_by(Country) %>% summarise(MissingYears = sum(is.na(Hepatitis.B)))
missingHepatitis<-missingHepatitis[order(missingHepatitis$MissingYears,decreasing=TRUE),]
missingHepatitis
#9 countries are missing data for all 16 years which only accounts for 144 of the missing values
#This is  different than population & GDP, may be able to impute majority from within rather than look for alternative sources


Hepatitis<-kaggle.training %>% select(-Country,-Status) %>% na.omit()

stat.test <- cor(Hepatitis[-7], Hepatitis$Hepatitis.B) 
stat.test
#most significant indivisual correlations:
#Diphtheria = 0.59 positive correlation makes since, higher vaccine rate for 1 disease = higher rate for most diseases
#Polio = 0.46 positive correlation makes since, higher vaccine rate for 1 disease = higher rate for most disease

Hepatitis %>% ggplot(aes(x = Diphtheria, y = Hepatitis.B)) + geom_point(size=0.5)

model <- lm(Hepatitis.B ~ ., data = Hepatitis)
summary(model)
#Adj R=0.41 , significant variables are Diphtheria, Polio, Year, infant.deaths, Measles, under.five.deaths, Schooling

model2 <- lm(Hepatitis.B ~ Diphtheria + Polio + Year + EstPopulation, data = Hepatitis)
summary(model2)
#Adj R=0.38 , significant variables are infant.deaths, Measles, under.five.deaths, Schooling

model3 <- lm(Hepatitis.B ~ Diphtheria + Polio, data = Hepatitis)
summary(model3)
#Adj R=0.36

model4 <- lm(Hepatitis.B ~ Diphtheria, data = Hepatitis)
summary(model4)
#Adj R=0.3475

Hepatitis["prediction"]<-predict(model, newdata = Hepatitis)

Hepatitis %>% ggplot(aes(x = Hepatitis.B, y = prediction)) + geom_point(size=0.5)
#Note that regression looks stronger for points with >50% rate of vaccination

#Look at sample countries with 7-9 missing years example Niger, Chad, Congo
Niger <- kaggle.training %>% filter(Country=="Niger") %>% select(Year,Hepatitis.B,Diphtheria) %>% gather(-Year, key = "Type", value = "value")
ggplot(Niger,aes(x = Year, y = value,color=Type)) + geom_line()

Chad <- kaggle.training %>% filter(Country=="Chad") %>% select(Year,Hepatitis.B,Diphtheria) %>% gather(-Year, key = "Type", value = "value")
ggplot(Chad,aes(x = Year, y = value,color=Type)) + geom_line()

Congo <- kaggle.training %>% filter(Country=="Congo") %>% select(Year,Hepatitis.B,Diphtheria) %>% gather(-Year, key = "Type", value = "value")
ggplot(Congo,aes(x = Year, y = value,color=Type)) + geom_line()
#wild swings in hepatitis rates by year, so replacing missing values with means by country may not be better than weak regression

#Check missing Hepatits data by Diptheria rates
missingHepatitis<-kaggle.training %>% group_by(Diphtheria) %>% summarise(MissingYears = sum(is.na(Hepatitis.B)))
missingHepatitis<-missingHepatitis[order(missingHepatitis$MissingYears,decreasing=TRUE),]
missingHepatitis
ggplot(missingHepatitis, aes(Diphtheria, MissingYears)) + geom_col()
#Note Majority of missing Hepatits points come from high Diphtheria rows, make stronger predictions just for those and leave other NA's rather than making bad predictions

Hepatitis2<-kaggle.training %>% select(-Country,-Status) %>% filter(Diphtheria>50) %>% na.omit()

model <- lm(Hepatitis.B ~ ., data = Hepatitis2)
summary(model)
#Adj R=0.41 , significant variables are Diphtheria, Polio, Year, infant.deaths, Measles, under.five.deaths, Schooling

model2 <- lm(Hepatitis.B ~ Diphtheria + Polio + Year + EstPopulation, data = Hepatitis2)
summary(model2)
#Adj R=0.38 , significant variables are infant.deaths, Measles, under.five.deaths, Schooling

model3 <- lm(Hepatitis.B ~ Diphtheria + Polio, data = Hepatitis2)
summary(model3)
#Adj R=0.36

```
Explore Correlations and relationships between variables

```{r corr-EDA}
#Create function to summarize the most significant correlations since there are too many variables for one plot
corr_simple <- function(data=df,sig=0.1){
  #convert data to numeric in order to run correlations
  #convert to factor first to keep the integrity of the data - each value will become a number rather than turn into NA
  df_cor <- data %>% mutate_if(is.character, as.factor)
  df_cor <- df_cor %>% mutate_if(is.factor, as.numeric)
  #run a correlation and drop the insignificant ones
  corr <- cor(df_cor)
  #prepare to drop duplicates and correlations of 1     
  corr[lower.tri(corr,diag=TRUE)] <- NA 
  #drop perfect correlations
  corr[corr == 1] <- NA 
  #turn into a 3-column table
  corr <- as.data.frame(as.table(corr))
  #remove the NA values from above 
  corr <- na.omit(corr) 
  #select significant values  
  corr <- subset(corr, abs(Freq) > sig) 
  #sort by highest correlation
  corr <- corr[order(-abs(corr$Freq)),] 
  #print table
  print(corr)
  #turn corr back into matrix in order to plot with corrplot
  mtx_corr <- reshape2::acast(corr, Var1~Var2, value.var="Freq")
  
  #plot correlations visually
  corrplot(mtx_corr, is.corr=FALSE, tl.col="black", na.label=" ")
}
corr_simple(kaggle.training)


numerical<-kaggle.training %>% select(-Country,-Status)
# ggpairs(numerical) to many combinations for plot to be useful

numerical %>%
  gather(-Life.expectancy, key = "var", value = "value") %>% 
  ggplot(aes(x = value, y = Life.expectancy, )) +
    geom_point(size=0.01) +
    facet_wrap(~ var, scales = "free") 

#Initial observations to explore:
#Adult.Mortality has high correlation appears to be two groups explore what separates groups
#Alcohol looks weak overall but may have clusters dependent on other factors
#BMI is similar
#Diptheria is similar
#GDP may need a transformation

#All should be explored with plotly to look at groupings
 #Why do so many variables have two distinct groupings in correlations? Is there a Country or data issue?

kaggle.training %>% ggplot(aes(x = Adult.Mortality, y = Life.expectancy)) +
    geom_point(size=0.5)

#Create manual grouping based on (0,90) to (150,40) which translates to Life = -0.33Mortality + 90
kaggle.training["Life.line"]=-0.33*kaggle.training$Adult.Mortality + 90

kaggle.training["grouping1"] <- ifelse(kaggle.training$Life.line >=kaggle.training$Life.expectancy, "A", "B")
kaggle.training$grouping1<-as.factor(kaggle.training$grouping1)

#verify manual grouping worked on scatterplot
kaggle.training %>% ggplot(aes(x = Adult.Mortality, y = Life.expectancy,color=grouping1)) +
    geom_point(size=0.5)

#Check to see if same grouping explains clustering on other scatter plots
numerical2<-kaggle.training %>% select(-Country,-Status) %>% gather(-Life.expectancy,-grouping1, key = "var", value = "value")
numerical2 %>% ggplot(aes(x = value, y = Life.expectancy,color=grouping1)) + geom_point(size=0.01) + facet_wrap(~ var, scales = "free") 
#Clusters on other scatters are not dependent on same manual groupings
```

For Manual grouping1 see what could explain difference between clusters

```{r}
stat.test <- numerical2 %>%
  group_by(var) %>%
  t_test(value ~ grouping1) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

#Sort and display numerical variables with highest significant difference by Attrition
stat.test <- stat.test %>% select(p, everything())
stat.test <- stat.test[order(stat.test$p),] 
stat.test

#Year is strongest correlation on grouping
kaggle.training %>% ggplot(aes(x=grouping1, y=Year)) + geom_boxplot()
kaggle.training %>% ggplot(aes(x=grouping1, y=GDP)) + geom_boxplot()
kaggle.training %>% ggplot(aes(x=grouping1, y=Measles)) + geom_boxplot()

kaggle.training %>% ggplot(aes(x = Adult.Mortality, y = Life.expectancy,color=Year)) +
    geom_point(size=0.5)

#No simple explanation for clustering

```

Try knn model on Life Expectancy vs. Mortality EXPLORE THIS MORE LATER

```{r}
knn.input<-na.omit(kaggle.training)

classifications = knn.cv(knn.input[,c(2,6:22)],knn.input$grouping1,prob = TRUE, k = 3)
confusionMatrix(table(classifications,knn.input$grouping1))

```

Plot variables vs population to see if some need to be normalized to per capita

```{r}
kaggle.training %>% ggplot(aes(x = Population, y = Adult.Mortality)) + geom_point(size=0.5)
kaggle.training %>% ggplot(aes(x = Population, y = infant.deaths)) + geom_point(size=0.5)
kaggle.training %>% ggplot(aes(x = Population, y = GDP)) + geom_point(size=0.5)

```